\documentclass[11pt,letterpaper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{indentfirst,amsmath,amssymb,enumitem}

\usepackage[utf8]{inputenc}
\usepackage[T2A]{fontenc}
\usepackage{textcomp}
\usepackage{DejaVuSansMono}

% https://tex.stackexchange.com/q/567250/35990
\newcommand{\textplus}[1][+]{\raisebox{% font-independent height
  \dimexpr(\fontcharht\font`X-\height+\depth)/2\relax
}{#1}}

\ExplSyntaxOn
\NewDocumentCommand\+{}{
  \peek_charcode_remove:NTF + {\textplus[++]}{\textplus[+]}
}
\ExplSyntaxOff

\usepackage{xcolor}

\definecolor{lstType}{HTML}{088A00}
\definecolor{lstKeyword}{HTML}{BF5908}
\definecolor{lstLiteral}{HTML}{990055}
\definecolor{lstNamespace}{HTML}{555555}
\definecolor{lstFunction}{HTML}{0077AA}
\definecolor{lstComment}{HTML}{708090}

\usepackage{listings}
\lstset{
    upquote=true,
%
    moredelim=[is][\color{lstType}]{@T}{@},
    moredelim=[is][\bfseries\color{lstType}]{@Tb}{@},
    moredelim=[is][\color{lstKeyword}]{@K}{@},
    moredelim=[is][\color{lstLiteral}]{@L}{@},
    moredelim=[is][\color{lstNamespace}]{@N}{@},
    moredelim=[is][\color{lstFunction}]{@F}{@},
%
    basicstyle={\footnotesize\ttfamily},
%
    morecomment=[l]{//}, % l is for line comment
    morecomment=[s]{/*}{*/}, % s is for start and end delimiter
    commentstyle={\itshape\color{lstComment}},
%
    morestring=[b]", % "string"
    morestring=[b]', % 'c'
    morestring=[s]{R"(}{)"}, % R"(raw string)"
    stringstyle=\color{lstLiteral},
    showstringspaces=false,
%
    morekeywords=[0]{
alignas, alignof, and, and_eq, asm, bitand, bitor, break, case, catch, class, co_await, co_return,
co_yield, compl, concept, const, const_cast, consteval, constexpr, constinit, continue,
contract_assert, decltype, default, delete, do, dynamic_cast, else, enum, explicit, export, extern,
for, friend, goto, if, inline, mutable, namespace, new, noexcept, not, not_eq, operator, or, or_eq,
private, protected, public, reflexpr, register, reinterpret_cast, requires, return, sizeof, static,
static_assert, static_cast, struct, switch, synchronized, template, this, thread_local, throw, try,
typedef, typeid, typename, union, using, virtual, volatile, while, xor, xor_eq
    },
    keywordstyle=[0]\color{lstKeyword},
%
    morekeywords=[1]{auto, void, char, int, short, long, signed, unsigned, float, double},
    keywordstyle=[1]\color{lstType},
%
    otherkeywords={::},
    keywords=[2]{::},
    keywordstyle=[2]\color{lstNamespace},
}

\usepackage[
  hyperfootnotes=false,
  linktoc=all,
  colorlinks=true,
  linkcolor=black,
  citecolor=black,
  urlcolor=blue,
  pdfauthor={Ivan Pogrebnyak},
  pdftitle={Units in C++},
  pdfsubject={},
  pdfkeywords={units, C++}
]{hyperref}

\renewcommand{\chapterautorefname}{Chapter}
\renewcommand{\sectionautorefname}{Section}
\let\subsectionautorefname\sectionautorefname
\let\subsubsectionautorefname\sectionautorefname

\newcommand{\lib}{\lstinline{@Nunits@} library}

\newcounter{lstcnt}[section]
\newenvironment{lstfloat}{%
  \refstepcounter{lstcnt}
  \noindent
  \begin{minipage}{0.9\textwidth}
}{%
  \end{minipage}\hfill(\thesection.\thelstcnt)
}

\title{Units in C\++}
\author{Ivan Pogrebnyak}
\date{\today}

\begin{document}

\maketitle

\section{Introduction} % ===========================================================================

Any physics or engineering simulation software has to account for units,
whether it is done explicitly or implicitly.
In this document, we discuss a new units library for C\++ that leverages the type system to
represent units.
The library provides features to define dimensionful quantities,
perform arithmetic operations on them, and validate compatibility of quantities at compile time.

\section{Theory} % =================================================================================

A physical \textbf{quantity}, such as volume or pressure, can be divided into two components:
\textbf{units}, specifying the type of quantity, and the quantity's \textbf{value}, specifying how
large the quantity is in the given units.

Geometrically, units can be viewed as \textbf{dimensions} of quantities. Much like a point
in 3-dimensional space can be represented by a vector $(x,y,z)$, quantities in mechanics have
dimensions of mass, length, and time. The units' dimensions are discrete-valued. For example,
represented as (mass, length, and time) vectors, speed units have dimensions $(0,1,-1)$, and those
for energy are $(1,2,-2)$.

Algebraically, units are multiplicative factors, orthogonal to the quantity's value.
For example, a length quantity can be expressed as 5 m, which can be thought of as multiplication
between a numeric value 5 and units m. Two lengths multiplied yields an area,
$5~\text{m} \times 3~\text{m} = 15~\text{m}^2$.
Note that both values and units were multiplied. Addition is typically only useful to define between
quantities with the same dimensions, while multiplication and division are meaningful in all cases.
For example, $\text{force} / \text{mass} = \text{acceleration}$.

Apples and oranges don't really behave algebraically like units. One is not typically interested in
oranges$^2$, while they may want to buy $5~\text{apples} + 3~\text{oranges}$.

Typically, multiple units exist for the same type of quantity, i.e. vector of dimensions. For
example, length can be measured in m, mm, km, mi, in, ft, etc. Mathematically, conversions between
these redundant units amount to multiplication by dimensionless factors. However, floating-point
arithmetic on a computer is not associative. When implementing a units system, it is important to
think about how conversions are handled and at what point. And, as always, one must find the optimal
trade-off between numeric precision, user friendliness, difficulty of implementation and
maintenance, and overhead at run time and compile time.

\section{The units library} % ======================================================================

\subsection{Example}\label{sec:example} % ---------------------------------------------------------

To illustrate what problems the \lib{} addresses and how it solves them,
let us start with a simple example.
Consider a physics simulation involving projectile motion.
For the sake of the argument, let's assume that there is a function that adds projectiles with a
given mass to the simulation.

\begin{lstlisting}
void @FAddProjectile@(double mass);
\end{lstlisting}

As is, the \lstinline{@FAddProjectile@()} function expects \lstinline{mass} in certain units.
These units are at best stated in a comment somewhere; and at worst, not stated anywhere at all.

It is not unreasonable for a library to internally assume a single mass unit, e.g. kg,
and perform all internal calculations without converting between kilograms and pounds.
But if \lstinline{@FAddProjectile@()} is a part of the interface, the value that should be passed to
it may come from another library that made a different assumption and chose pounds as base units
instead of kilograms. The situation may be as follows.
%
\begin{lstlisting}
double @Nfoo@::@FGetMass@(); // returns mass in lb
void @Nbar@::@FAddProjectile@(double mass); // takes mass in kg
\end{lstlisting}
%
A call like this would be incorrect.
%
\begin{lstlisting}
@Nbar@::@FAddProjectile@(@Nfoo@::@FGetMass@()); // Mistakenly forgot to convert lb to kg
\end{lstlisting}
%
A software engineer who read the documentation for both libraries is the only one that can catch
this mistake.
The compiler cannot provide any help here.

The \lib{} can help with this problem.
Exactly what is the best approach, depends on whether we own
\lstinline{@Nfoo@} or \lstinline{@Nbar@}.

First, let's assume that we can modify neither
\lstinline{@Nfoo@} nor \lstinline{@Nbar@}.
In this case, the best we can do is apply a conversion factor.
It may be tempting to say that in this case a library does not have much to offer, as we can just
define a conversion factor.
%
\begin{lstlisting}
const double lb_to_kg = @L0.45359237@;
@Nbar@::@FAddProjectile@(@Nfoo@::@FGetMass@() * lb_to_kg);
\end{lstlisting}
%
But how many conversion factors does a whole simulation application codebase need?
How do we avoid redefining the same conversion factors?
How do we make sure the factors are defined accurately and consistently?
Using the \lib, we can avoid these questions and use consistent definitions.
%
\begin{lstlisting}
using namespace @Nunits@::@Nliterals@;
@Nbar@::@FAddProjectile@(@Nfoo@::@FGetMass@() * "kg/lb"@L_u@);
\end{lstlisting}

If we own \lstinline{@Nbar@} but not \lstinline{@Nfoo@}, we can change the signature of
\lstinline{@FAddProjectile@()} to take a mass quantity.
%
\begin{lstlisting}
using @Nunits@::@Nquantities@::@TMass@;
void @Nbar@::@FAddProjectile@(@TMass@<double> mass); // units-aware definition
\end{lstlisting}
%
With this change,
%
\begin{lstlisting}
@Nbar@::@FAddProjectile@(@Nfoo@::@FGetMass@()); // Compilation error: dimensionless
@Nbar@::@FAddProjectile@(@L5@); // Compilation error: dimensionless
@Nbar@::@FAddProjectile@(@L7.1@ * "m"@L_u@); // Compilation error: meters are not mass units
\end{lstlisting}
%
all produce errors at compile-time, as neither of the arguments are convertible to
\lstinline{@TMass@<double>}.
Valid calls would look like this.
%
\begin{lstlisting}
@Nbar@::@FAddProjectile@(@Nfoo@::@FGetMass@() * "lb"@L_u@);
@Nbar@::@FAddProjectile@(@L7.1@ * "kg"@L_u@);
\end{lstlisting}
%
We multiply by \lstinline{"lb"@L_u@} to specify the units of the quantity that we have.
The \lib{} uses its definition of pounds to provide the right mass quantity.
The code becomes self-documenting and unit correctness is checked at compile time.

If we own \lstinline{@Nfoo@} but not \lstinline{@Nbar@}, we can change the signature of
\lstinline{@FGetMass@()} to return a mass quantity.
%
\begin{lstlisting}
using @Nunits@::@Nquantities@::@TMass@;
@TMass@<double> @Nfoo@::@FGetMass@(); // units-aware definition
\end{lstlisting}
%
The corresponding call to unmodified \lstinline{@FAddProjectile@()} would look like this,
%
\begin{lstlisting}
@Nbar@::@FAddProjectile@(@Nfoo@::@FGetMass@() / "kg"@L_u@);
\end{lstlisting}
%
We divide by \lstinline{"kg"@L_u@} because we need the value in kilograms.
Here, again, the code becomes self-documenting, as units have to be explicitly stated, or it won't
compile.

And finally, if we own both \lstinline{@Nfoo@} and \lstinline{@Nbar@}, we can modify both functions
to operate on units-aware quantities, making the following expression valid and correct.
%
\begin{lstlisting}
@Nbar@::@FAddProjectile@(@Nfoo@::@FGetMass@());
\end{lstlisting}
Note, that now, if the above expression compiles, we have a strong guarantee that the quantities
are compatible between the two functions.

\subsection{Implementation synopsis} % -------------------------------------------------------------

The main construct of the \lib{} is
%
\begin{lstlisting}
template <@TDimensions@ D, typename @TT@>
class @TQuantity@ {
    // . . .
};
\end{lstlisting}
%
\lstinline{@TQuantity@} wraps a value of type \lstinline{@TT@} and tags it with units of dimensions
\lstinline{D}.
The \lstinline{typename @TT@} template parameter is constrained to not be a \lstinline{@TQuantity@}
to avoid constructs with multiple levels of units.
%
\begin{lstlisting}
struct @TDimensions@ {
    int mass, length, time;
    // . . .
};
\end{lstlisting}
%
implements a \lstinline{constexpr} integral-valued vector with arithmetic operations
for multiplication and exponentiation of quantities.
The tagging is done purely to introduce units to the type system and has no effect at run time.
At compile time, \lstinline{@TDimensions@} provide means to check compatibility of quantities'
units and units' arithmetic.

The library provides a number of predefined aliases for specific quantity types.
For example, speed and velocity are represented as
%
\begin{lstlisting}
template <typename @TT@>
using @TSpeed@ = @TQuantity@<{@L0@, @L1@, @L-1@}, @TT@>;
template <typename @TT@>
using @TVelocity@ = @TSpeed@<@TT@>;
\end{lstlisting}

To reduce the space of possible mistakes and to keep the implementation details of the library
opaque to the user, the \lstinline{@TQuantity@} constructor is made private,
except for the dimensionless case.
This prevents the user from creating dimensionful quantities with ambiguous units.
%
\begin{lstlisting}
@TQuantity@<{@L1@, @L0@, @L0@}, double> mass(@L5.3@); // Compilation error
                                       // What are the units of this mass?
@TQuantity@<{@L0@, @L0@, @L0@}, double> number(@L7.13@); // OK
                                          // This is a dimensionless number
\end{lstlisting}

The intended mechanism of creating dimensionful quantities in the \lib{} is via
user-defined literals.
A literal constructs a \lstinline{constexpr} \lstinline{@TQuantity@} with the value equal to
$1$ in the corresponding units.
The literals are defined under the \lstinline{@Nunits@::@Nliterals@} namespace.
For example,
%
\begin{lstlisting}
using namespace @Nunits@::@Nliterals@;
auto length = "m"@L_u@; // equivalent to 1 meter
\end{lstlisting}
%
The type of \lstinline{length} is \lstinline|@TQuantity@<{@L0@, @L1@, @L0@}, double>|.

The underscore in the \lstinline{@L_u@} suffix is required by the C\++ standard,
because literal suffixes without a leading underscore are reserved for the standard library.
The \lstinline{@L_u@} suffix creates \lstinline{double} quantities.
The library also provides a \lstinline{@L_uf@} suffix to create \lstinline{float} quantities.

The intended way to use the \lstinline{@Nunits@} literals is to specify both the value and
the units via multiplication.
%
\begin{lstlisting}
auto length = @L2.3@ * "m"@L_u@; // equivalent to 2.3 meter
\end{lstlisting}
%
Thus, multiplication by a literal can be thought of as a way to imbue values with units.
Conversely, division by a literal can be used to convert a quantity to specific units.
%
\begin{lstlisting}
auto length = @L3@ * "ft"@L_u@;
double length_in_ft = length / "m"@L_u@; // equal to 0.9144
\end{lstlisting}

More generally, division and multiplication are implemented for all
\lstinline{@TQuantity@<D, @TT@>} for which the value type \lstinline{@TT@} has
its division and multiplication implemented respectively.
Addition and subtraction are implemented likewise, but only for pairs of
quantities with equal dimensions.

Dimensionless quantities are implicitly convertible to references to their values.
This feature works in tandem with multiplication and division operators.
For example, assuming that \lstinline{@TVec3@} implements a 3-dimensional vector type
with overloaded operators.
%
\begin{lstlisting}
using namespace @Nunits@::@Nquantities@;
@TLength@<@TVec3@> @FGetDisplacement@();
@TTime@<double> @FGetTime@();
auto velocity = @FGetDisplacement@() / @FGetTime@(); // the type is Velocity<Vec3>
\end{lstlisting}

To further illustrate how these features work together,
consider a situation where we have a function that we can't modify that takes velocity
in $\text{m}/\text{s}$ as \lstinline{@TVec3@}.
%
\begin{lstlisting}
void @Nthirdparty@::@FCalculate@(const @TVec3@&);
\end{lstlisting}
%
As in the example in \autoref{sec:example}, we can call this function like this
%
\begin{lstlisting}
@Nthirdparty@::@FCalculate@(velocity / "m/s"@L_u@);
\end{lstlisting}
%
This works, because the result of division by the literal is a dimensionless quantity,
and as such, it is convertible to a reference to it's value.
The following, however, do not work.
%
\begin{lstlisting}
@Nthirdparty@::@FCalculate@(velocity); // compilation error
@Nthirdparty@::@FCalculate@(position / "m/s"@L_u@); // compilation error
\end{lstlisting}
%
In both cases, the quantities passed to the function are not dimensionless.

This combination of features provides some good guarantees. However, it relies on the
function call requiring an implicit conversion.
The following does not produce a compilation error, though there is a bug.
%
\begin{lstlisting}
auto velocity_in_mph = position / "mph"@L_u@; // compiles, but is likely a bug
\end{lstlisting}

To provide a way to get the value of a quantity in specific units and simultaneously check
that the dimensions are compatible, the \lib{} implements the
\lstinline{@FTo@()} function.
\begin{lstlisting}
auto not_velocity_in_mph = @Nunits@::@FTo@<"mph">(position); // compilation error
auto velocity_in_mph = @Nunits@::@FTo@<"mph">(velocity); // OK
\end{lstlisting}

\section{Implementation choices} % =================================================================

\subsection{Run-time or compile-time representation of units} % ------------------------------------

\subsection{Immediate or delayed conversion} % -----------------------------------------------------

\subsection{Static or dynamic number of dimensions} % ----------------------------------------------

\subsection{Distinguishing units with equal dimensions but distinct meaning} % ---------------------

Dimensionless quantities: degrees, radians. Degrees as a factor: std::sin(15 * "deg"\_u).

\section{Future features} % ========================================================================

Features that can but haven't been implemented.

\subsection{Rational powers} % ---------------------------------------------------------------------

\subsection{User-defined units and relative definitions} % -----------------------------------------

\InputIfFileExists{extra}{}{}

\end{document}
